Testing gene methylation enrichment analysis approaches using simulated data

### 2023-11-17 figs/GSE158422_simulate020.Rmd
forece use of epic chip annotation

simgsa <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
  # generate gene sets
  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
  lengths <- unname(unlist(lapply(genesetdatabase,length)))
  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
  # select gene sets to alter
  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
  # incorporate select changes
  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta)
  # set up limma
  mval2 <- sim$mval
  ncols <- ncol(mval2)
  groupsize <- ncols/2
  ss <- data.frame(colnames(mval2))
  colnames(ss) <- "sample"
  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
  d <- model.matrix(~ ss$case )
  dm3 <- runlimma(mval=mval2,design=d,myann=myann)
  pup3 <- rownames(subset(dm3,adj.P.Val<0.05 & logFC>0))
  pdn3 <- rownames(subset(dm3,adj.P.Val<0.05 & logFC<0))
  if ( length(pup3) < 250 ) { pup3 <- head(rownames(subset(dm3, logFC > 0)), 250) }
  if ( length(pdn3) < 250 ) { pdn3 <- head(rownames(subset(dm3, logFC < 0)), 250) }
  # convert gene sets to entrez
  suppressWarnings(suppressMessages({ gene2entrez <- mapIds(org.Hs.eg.db, gene_catalog, 'ENTREZID', 'SYMBOL') }))
  gsets_entrez <- lapply(gsets,function(gs) {
    gs2 <- unique(gene2entrez[names(gene2entrez) %in% gs])
    gs2 <- gs2[!is.na(gs2)]
    return(gs2)
  })
  suppressWarnings(suppressMessages({
    gsaup3 <- gsameth(sig.cpg=pup3, all.cpg=rownames(dm3), collection=gsets_entrez)
    gsadn3 <- gsameth(sig.cpg=pdn3, all.cpg=rownames(dm3), collection=gsets_entrez)
  }))
  gsig_up3 <- rownames(subset(gsaup3,FDR<0.05))
  gsig_dn3 <- rownames(subset(gsadn3,FDR<0.05))
  gtup <- names(sim[[6]])
  gtdn <- names(sim[[7]])
  UPTP=length(intersect(gsig_up3 ,gtup))
  UPFP=length(setdiff(gsig_up3 ,gtup))
  UPFN=length(setdiff(gtup,gsig_up3))
  DNTP=length(intersect(gsig_dn3 ,gtdn))
  DNFP=length(setdiff(gsig_dn3 ,gtdn))
  DNFN=length(setdiff(gtdn,gsig_dn3))
  TP=UPTP+DNTP
  FP=UPFP+DNFP
  FN=UPFN+DNFN
  TN=nrow(gsadn3)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
  PREC=TP/(TP+FP)
  REC=TP/(TP+FN)
  F1=TP/(TP+(0.5*(FP+FN)))
  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
  return(result)
}

>>>>

simgsa <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
  # generate gene sets
  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
  lengths <- unname(unlist(lapply(genesetdatabase,length)))
  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
  # select gene sets to alter
  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
  # incorporate select changes
  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta)
  # set up limma
  mval2 <- sim$mval
  ncols <- ncol(mval2)
  groupsize <- ncols/2
  ss <- data.frame(colnames(mval2))
  colnames(ss) <- "sample"
  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
  d <- model.matrix(~ ss$case )
  dm3 <- runlimma(mval=mval2,design=d,myann=myann)
  pup3 <- rownames(subset(dm3,adj.P.Val<0.05 & logFC>0))
  pdn3 <- rownames(subset(dm3,adj.P.Val<0.05 & logFC<0))
  if ( length(pup3) < 250 ) { pup3 <- head(rownames(subset(dm3, logFC > 0)), 250) }
  if ( length(pdn3) < 250 ) { pdn3 <- head(rownames(subset(dm3, logFC < 0)), 250) }
  # convert gene sets to entrez
  suppressWarnings(suppressMessages({ gene2entrez <- mapIds(org.Hs.eg.db, gene_catalog, 'ENTREZID', 'SYMBOL') }))
  gsets_entrez <- lapply(gsets,function(gs) {
    gs2 <- unique(gene2entrez[names(gene2entrez) %in% gs])
    gs2 <- gs2[!is.na(gs2)]
    return(gs2)
  })
  suppressWarnings(suppressMessages({
    gsaup3 <- gsameth(sig.cpg=pup3, all.cpg=rownames(dm3), collection=gsets_entrez, array.type="EPIC")
    gsadn3 <- gsameth(sig.cpg=pdn3, all.cpg=rownames(dm3), collection=gsets_entrez, array.type="EPIC")
  }))
  gsig_up3 <- rownames(subset(gsaup3,FDR<0.05))
  gsig_dn3 <- rownames(subset(gsadn3,FDR<0.05))
  gtup <- names(sim[[6]])
  gtdn <- names(sim[[7]])
  UPTP=length(intersect(gsig_up3 ,gtup))
  UPFP=length(setdiff(gsig_up3 ,gtup))
  UPFN=length(setdiff(gtup,gsig_up3))
  DNTP=length(intersect(gsig_dn3 ,gtdn))
  DNFP=length(setdiff(gsig_dn3 ,gtdn))
  DNFN=length(setdiff(gtdn,gsig_dn3))
  TP=UPTP+DNTP
  FP=UPFP+DNFP
  FN=UPFN+DNFN
  TN=nrow(gsadn3)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
  PREC=TP/(TP+FP)
  REC=TP/(TP+FN)
  F1=TP/(TP+(0.5*(FP+FN)))
  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
  return(result)
}
